---
title: "Predicting Citi Bike Usage in NYC"
author: "David Skålid Amundsen"
date: "March 15, 2016"
output: beamer_presentation
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
# Load packages and data frames
library(dplyr)
library(lubridate)

load("data/data_frames.dat")
load("data/models.dat")
```

## Goal: Predict Citi Bike Usage in NYC

- Demand varies with time, but what determines it?

- What is demand?
    - Average trip length.
    - Number of bikes on the road.
    - Number of new bike rentals per time.
    
- I will try to predict:
    - Number of new bike rentals per time.
    - Predictors: **time** and **weather**.

## Citi Bike Data

Details each trip:

- Duration
- Start and end time
- Stations
- Age
- Gender
- Bike ID.

I reduce it to:

- Time
- Number of new bike rentals

## Citi Bike Usage on July 4th 2015

```{r, echo = FALSE}
library(ggplot2)

df_plot <- df_citibike %>%
  filter(year(time) == 2014 & month(time) == 4 & (day(time) >= 3 & day(time) <= 5))
ggplot(data = df_plot, aes(x = time, y = numtrips)) +
  geom_bar(stat = "identity") +
  xlab("Time") +
  ylab("Number of new rentals")
```

## Weather Data

Weather observations from NOAA's weather station in central park:

- Wind speed
- Direction
- Temperature
- Precipitation
- Pressure
- Dew point

## Temperature observations

```{r, echo = FALSE}
library(ggplot2)

df_plot <- df_weather %>%
  filter(year(time) == 2014 & month(time) == 4 & (day(time) >= 3 & day(time) <= 5))
ggplot(data = df_plot, aes(x = time, y = TEMP)) +
  geom_bar(stat = "identity") +
  xlab("Time") +
  ylab("Temperature [degrees C]")
```

## Data for modelling

Using the following predictors:

- Weather:
    - Wind speed and direction
    - Temperature
    - Precipitation
    - Pressure
    - Dew point.
- Time:
    - Hour of day
    - Weekday

Consequence:

- Any seasonal variations are modelled by weather.

## Model result:

Random forest provided best fit and prediction:

- RMSE: 360 trips per hour
- R$^2$: 0.85
- RMSE on test set: 362 trips per hour

## Test prediction on random day in April 2015 (13th)

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(gridExtra)

pr_day_rf <- predict(rf_fit, df_day_test)

# Check interpolation
plot_data <- ggplot() +
  geom_bar(aes(x = df_day_test$hour, y = df_day_test$numtrips), stat = "identity") +
  ylim(0, 3500) +
  xlab("Hour of day") +
  ylab("Actual number of trips")

plot_predict <- ggplot() +
  geom_bar(aes(x = df_day_test$hour, y = pr_day_rf), stat = "identity") +
  ylim(0, 3500) +
  xlab("Hour of day") +
  ylab("Predicted number of trips")

grid.arrange(plot_data, plot_predict, ncol = 2)
```

## Further improvments

- Depends on major events?
- Yearly trends?
- Depends on station location?
- Depends on cloud cover?

## Other models applied

Generalized linear model:

- RMSE of fit: 497 trips per hour
- $R^2$: 0.72
- RMSE of test set: 506 trips per hour

Regression tree:

- RMSE of fit: 419 trips per hour
- $R^2$: 0.80
- RMSE of test set: 414 trips per hour